---
title: 'Clustering'
author: ""
date: ""
output:
  html_document:
    toc: true
    toc_float: true
---


```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=7, fig.height=5) 
```

## Goal

In this lab we will learn the basics of clustering. The methods we will cover 
include hierarchical clustering, K means and density clustering.

## Packages

Install and load packages.

```{r setup, warning=FALSE, message=FALSE, results='hide'}
pkgs_needed = c(
  "tidyverse",
  "dbscan", "GGally", "pheatmap",
  "flowCore","flowViz","flowPeaks", "ggcyto"
)
letsinstall = setdiff(pkgs_needed, installed.packages()) 
if (length(letsinstall) > 0) {
  for (pkg in letsinstall)
    BiocManager::install(pkg)
}
```

```{r, warning=FALSE, message=FALSE, results='hide'}
library("tidyverse")
```

## Hierarchical clustering

The Morder data are gene expression measurements for 156 genes on T cells of 
3 types (naïve, effector, memory) from 10 patients (Holmes et al. 2005).
Here we load the `Morder` data.frame from the online directory.

```{r}
load(url("http://web.stanford.edu/class/bios221/data/Morder.RData"))
dim(Morder)
```

In 'base' R a function to perform hierarchical clustering is `hclust()`.
To cluster the genes with hierarchical clustering you first need
to compute a distance matrix storing all pairwise (gene-to-gene)
dissimilarities. The following commands would be useful:

```{r, fig.width=15, fig.height=8}
# distance calculation
D <- dist(t(Morder))

# clustering
gene_clust <- hclust(d = D)

# plot dendogram
plot(gene_clust)
```

**Quiz question 6**: Why in the provided code the input to `dist()`
function is `t(Morder)`? 

**Answer**: 

In hierarchical clustering one needs to choose the method for agglomerating the 
clusters. By default `hclust()` uses a "complete" linkage method. Please redo 
hierarchical clustering with "ward.D2" method. Note that in the `hclust()` there 
are "ward.D" and "ward.D2" methods available. Please call `?hclust` to read
about the difference between the two methods.

```{r}
gene_clust <- hclust(d = dist(t(Morder)), method = "ward.D2")
```

**Quiz question 7**: Note that the height of the dendrogram is changed when
you redid the clustering with a different linkage method. What do the y-axis 
values on the hclust dendrogram plot correspond to?

**Answer:** distance between clusters (according to agglomeration method choice)

Now, instead of clustering genes, apply hierarchical clustering
for samples (observations), with default linkage method.

```{r}
# we don't transpose the matrix now (samples are rows)
D_samples <- dist(Morder)
sample_clust <- hclust(d = D_samples)
plot(sample_clust)
```

**Quiz question 8**: How many clusters of samples are there at the dendrogram 
height of 12. Hint the `abline()` function might be helpful.


**Answer**: 

```{r}
plot(sample_clust)
abline(a=12, b=0, col="blue")
```

Now that you know how to perform hierarchical clustering, use `pheatmap()`
to generate a heatmap with rows and columns grouped according
to computed dendrograms. 

```{r fig.height=10, fig.width=8}
library(pheatmap)
pheatmap(Morder, fontsize_col=12, fontsize_row = 15) 
```


**Quiz question 9**: In `pheatmap` you can specify what distance metric to 
compute for clustering rows and columns. What type of distance does `pheatmap`
use by default?

**Answer**: 

**Quiz question 10**: What type of clustering (agglomeration) method does
`pheatmap` use by default?

**Answer**: 

## Density based clustering
### Mass cytometry (CyTOF)

Cytometry is a biophysical technology that allows you to measure
physical and chemical characteristics of cells. Modern flow and mass 
cytometry allows for simultaneous multiparametric analysis of thousands of 
particles per second.

[Flow cytometry](https://en.wikipedia.org/wiki/Flow_cytometry) enables the 
simultaneous measurement of 15, whereas 
[mass cytometry (CyTOF)](https://en.wikipedia.org/wiki/Mass_cytometry) of as 
many as 40 proteins per single cell.

We start by downloading and reading in a CyTOF dataset. The dataset 
comes from a single-cell mass cytometry study by 
[Bendall et al.](http://science.sciencemag.org/content/332/6030/687.full)
on differential immune drug responses across human hematopoietic cells
over time.

We first download and then load this data in R using flowCore package.

```{r}
library("flowCore")
library("flowViz")
library("flowPeaks")
library("ggcyto")

# download file from web (14.3 Mb)
download.file(
  url = "http://web.stanford.edu/class/bios221/data/Bendall_2011.fcs",
  destfile = "Bendall_2011.fcs",
  mode = "wb"
)

# if download fails with timeout error, try increasing it
# options(timeout = 1000)

# load file in R - here we use a function from flowCore package
fcsB = flowCore::read.FCS("Bendall_2011.fcs")
fcsB

# inspect slots of the flowFrame object
slotNames(fcsB)
```

Different slots of the `flowFrame` objects can be accessed with `exprs(fcsB)`, `parameters(fcsB)` and `description(fcsB)` functions.

**Quiz question 1**: Look at the structure of the `fcsB` object. How many variables were measured? 

**Answer**:

```{r}
length(colnames(fcsB))
```

**Quiz question 2**: How many cells were measured in the `fcsB` object?

**Answer**:

```{r}
nrow(exprs(fcsB))
```

### Data preprocessing

First we load the data table that reports the mapping between isotopes and 
markers (antibodies); then, we replace the isotope names in the column 
names of `fcsB` with the marker names. This is simply to make the subsequent 
analysis and plotting code more readable.

```{r}
# download data from the web
markersB = read_csv(url("http://web.stanford.edu/class/bios221/data/Bendall_2011_markers.csv"))

# match isotope names to marker names
mt = match(markersB$isotope, colnames(fcsB))

# check that all isotope names were matched
stopifnot(!any(is.na(mt)))

# replace isotope names with marker names
colnames(fcsB)[mt] = markersB$marker
```

Below, we show how to plot the joint distribution of the cell lengths and 
the `DNA191` (indicates the activity of the cell whether cell is dead or alive).
The information is included in the `fcsB` object.

```{r}
flowPlot(fcsB, plotParameters = c("Cell_length", "DNA191"), logy=TRUE)
```

It is standard to transform both flow and mass cytometry data using one of 
several special functions, we take the example of the inverse hyperbolic sine 
(arcsinh), which serves as a variance stabilizing transformation. 
First, we show the distribution on untransformed raw data:

```{r}
densityplot(~`CD3all`, fcsB)
```

To apply the transformation and to plot the data you can use functions
from the `flowCore`` package. After the transformation the cells seem to 
form two clusters: Based solely on one dimension (CD3all) we see two cell subsets 
(the two modes).

```{r}
asinhT = arcsinhTransform(a = 0.1, b = 1)
cols_to_transform <- setdiff(colnames(fcsB), c("Time", "Cell_length", "absoluteEventNumber"))
trans1 = transformList(cols_to_transform, asinhT)
fcsBT = transform(fcsB, trans1)
densityplot(~`CD3all`, fcsBT)
```

### K-means

Let's cluster cells into two groups using one-dimensional k-means filter.
To learn more about the arguments of the functions type `?kmeansFilter`
and `?flowCore::filter`

```{r}
kf = kmeansFilter("CD3all"=c("Pop1","Pop2"), filterId="myKmFilter")
fres = filter(fcsBT, kf)
summary(fres)
```

Split two data clusters into two separate flowFrame objects:

```{r}
fcsBT1 = split(fcsBT, fres, population="Pop1")
fcsBT2 = split(fcsBT, fres, population="Pop2")
```

We can also cluster cell with the `flowPeaks()` function from `flowPeaks` 
package. The algorithm for this clustering algorithm is specified
in detail in a paper by
[Ge Y. et al (2012)](https://academic.oup.com/bioinformatics/article/28/15/2052/236255).


```{r}
dat = data.frame(exprs(fcsBT)[ , c("CD56", "CD3all")])
fp = flowPeaks(dat)
plot(fp)
```

**Quiz question 3**: How many dimensions (markers) does the above code use 
to split the data into 4 cell subsets using kmeans?

**Answer:** 2 dimensions, based on `CD56` and `CD3all`

A Bioconductor package ``ggcyto`` build on top of ``ggplot2`` 
includes functions for generating visualizations specifically
for cytometry data. Note that here `fcsB` or `fcsBT` are not 
'data.frames' but objects of class  'flowFrame'. This means
that you cannot use `fcsB` and `fcsBT`  (without conversion to data.frame)
as inputs to `ggplot()`. 'flowFrame' objects hold marker expression
data and sample information data, so you can access any variables you need.

```{r ggcytoCD4CD8}
library("ggcyto")
# Untransformed data
ggcyto(fcsB,aes(x = CD4)) + geom_histogram(bins = 60) 
```

```{r}
# Transformed data
ggcyto(fcsBT, aes(x=CD4)) + geom_histogram(bins=90) 
```

```{r}
# ggcyto automatic plotting
autoplot(fcsBT, "CD4")
```

```{r}
ggcyto(fcsBT, aes(x = CD4, y = CD8)) + geom_density2d(colour="black") 
```

```{r}
ggcyto(fcsBT, aes(x = CD4, y = CD8)) + geom_hex(bins = 50) 
```


```{r}
# ggcyto automatic plotting
autoplot(fcsBT, "CD4", "CD8", bins = 50)
```

For more details on capabilities of `ggcyto` refer to the following
[link](https://bioconductor.org/packages/release/bioc/vignettes/ggcyto/inst/doc/autoplot.html)

### Density based clustering

Data sets such as flow cytometry containing only a few markers and a large 
number of cells are amenable to density clustering. DBSCAN algorithm looks for 
regions of high density separated by sparse emptier regions. This method has 
the advantage of being able to cope with clusters that are not necessarily 
convex (i.e. not blob-shaped). One implementation of such a method is called 
``dbscan``, let us look at an example by running the code below:

```{r dbscanfcs5}
library("dbscan")

# Select a small subset of 5 protein markers
mc5 = exprs(fcsBT)[, c("CD4", "CD8", "CD20", "CD3all", "CD56")]
res5 = dbscan::dbscan(mc5, eps = .65, minPts = 30)
mc5df = data.frame(mc5)
mc5df$cluster = as.factor(res5$cluster)
```


**Quiz question 4**: How many clusters did `dbscan()` find?

**Answer:** 
7 clusters, as we can see by counting all clusters that were found.

```{r}
table(mc5df$cluster)
```

Note that we do not count the `0` as a cluster, since it contains all cells that were not assigned to any cluster. Looking up the documentation  of `dbscan` (`?dbscan`) we find that "Points which cannot be assigned to a cluster will be reported as members of the noise cluster 0."

**Quiz question 5**: How many cells were clustered into cluster 3
by `dbscan()`?

**Answer**: 

```{r}
sum(mc5df$cluster == 3)
```

We can now generate a CD8-vs-CD4 2d-density plot for the cells colored
by their assigned cluster labels, computed by `dbscan()`:

```{r}
ggplot(mc5df, aes(x = CD4, y = CD8, colour = cluster)) + geom_density2d()
```

Adn do the same for CD3all and CD20 markers:

```{r}
ggplot(mc5df,aes(x = CD3all, y = CD20, colour = cluster))+ geom_density2d()
```

Observe that **the nature nature of the clustering is multidimensional**,
as the projections into two dimensions show overlapping clusters.

## Validating and choosing the number of clusters

The clustering methods we have described are tailored to deliver the best 
grouping of the data under various constrains, however they will always deliver 
groups, even if there are none. This is important, e.g. when performing
kmeans clustering, as we have to set the 'k' parameter (for the number
of clusters to group observations into) ahead of time. What choice of 'k'
is valid though?

Here we want to illustate the use of the "wss" (within sum of squares) statistic 
to evaluate the quality of a clustering. Note that as $k$ (number of cluster for 
k-means algorithm) increases, wss will also decrease. We simulate data coming 
from 4 groups. In particular, we generate 2-dimensional observations
(as if there were only 2 proteins measured for each cell). The four groups
are generated from 2-d multivariate normals with centers at
$\mu_1 = (0, 0)$, $\mu_2 = (0, 8)$, $\mu_3 = (8, 0)$, $\mu_4 = (8, 8)$.
In this simulation, we know the ground truth (4 groups), but we will
try to cluster the data using the `kmeans` argorithm with different choices for the 'k'
parameter. We will see how the wss statistic varies as we vary `k`.

We have  used the `%>%` operator from the ``dplyr`` package (if you do not 
understand the code, try to see what `simul4` contains and repeat the same using
code that does not use the `%>%` operator).

```{r Fake4}
simul4 = lapply(c(0,8), function(x){
  lapply(c(0,8), function(y){
    data.frame(x = rnorm(100, x, 2),
               y = rnorm(100, y, 2), 
               class = paste(x, y, sep = "")
    )
  }) %>% do.call(rbind,.)
}) %>% do.call(rbind,.)
```

```{r}
ggplot(simul4, aes(x = x, y = y)) +
  geom_point(aes(color = class), size = 2)
```

```{r}
# Compute the kmeans within group wss for k=1 to 12
wss = rep(0,8)
# for a single cluster the WSS statistic is just sum of squares of centered data
wss[1] = sum(apply(scale(simul4[,1:2], scale = F), 2, function(x){ x^2 }))
# for k = 2, 3, ... we perform kmeans clustering and compute the associated WSS statistic
for (k in 2:8) {
  km4 <- kmeans(simul4[,1:2],k)
    wss[k] =  sum(km4$withinss)
}
# Now, we are ready to plot the computed statistic:
ggplot(data.frame(k = 1:length(wss), wss = wss)) +
  geom_point(aes(x = k, y = wss), color = "blue", size= 3) +
  xlab('k') + ylab('WSS(k)')
```

Within sum of squares (wss) statistic, we see that the last substantial decrease
of the statistic occurres before $k=4$, and for values $k = 5, 6, \dots$
the quantity 'levels-off'. In practice, we would choose $k=4$, a value
happening at the 'elbow' of the plot (elbow-rul). Of course this choice is 
still somewhat subjective. The book chapter describes additional ways of
choosing `k` (e.g. the gap statistic).
